A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-27 tex/BiSig.tex
[ tex ] Minor changes in S1
[ tex ] Done with S1 for now

\setlength {\marginparwidth }{2cm}
\usepackage[color=yellow,textsize=small]{todonotes}

@@

  In response to this difficulty, we develop a simple yet general and constructive theory of bidirectional type synthesis where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing rules to determine the type of a program with type annotations.
  First, our theory is simple yet general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
  Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated positively in order to compute.
  Specifically, the `proof' of logical decidability of type synthesis is instantiated to a synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.
  We introduce the notion of mode preprocessing to refine annotatability in bidirectional typing.
  Type synthesis and mode preprocessing result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception and locate missing annotations.
  By formalizing our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.
\end{abstract}

@@

\keywords{\Agda, abstract syntax, bidirectional typing, type synthesis, type checking, type inference, language formalisation, datatype-generic programming}

>>>>

\setlength {\marginparwidth }{2cm}
\usepackage[obeyFinal,color=yellow,textsize=small]{todonotes}

@@

  In response to this difficulty, we develop a simple yet general and constructive theory of bidirectional type synthesis where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing rules to determine the type of a program with type annotations.
  First, our theory is simple yet general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
  Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated positively in order to compute.
  Specifically, the `proof' of logical decidability of type synthesis is instantiated to a synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.
  We introduce the notion of mode preprocessing to refine annotatability in bidirectional typing.
  Type synthesis and mode preprocessing result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception and locate missing annotations.
  By formalising our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.
\end{abstract}

@@

\keywords{bidirectional typing, type synthesis, type checking, type inference, language formalisation, \Agda}

### 2023-06-27 tex/sec1-intro.tex
[ tex ] Done with S1 for now

Type checking and inference serve as the transition to type-checked programs, also known as \emph{well-typed terms}, from parsed abstract syntax trees, referred as \emph{raw terms} by contrast.
Type inference algorithms were conceived to ascertain the type of any raw term without any type annotations.
However, it was later found that full parametric polymorphism leads to undecidability in type inference, as do dependent types~\citep{Wells1999,Dowek1993}.
In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing decidable algorithms for determining the types of \emph{suitably annotated} programs in languages with more complex types. 
\citet{Dunfield2021} summarised its design principles and applications that we know so far.

The idea of bidirectional type synthesis begins with extending typing judgements with two modes that direct the flow of types:
\begin{enumerate*}
  \item $\Gamma |- \isTerm{t} \syn A$ for synthesis and 
  \item $\Gamma |- \isTerm{t} \chk A$ for checking.
\end{enumerate*}
The former means that the type of a raw term is synthesised, using both the term and its context as input, while for the latter a raw term is to be checked against a given type.
Then a type synthesis algorithm can be `read off' from a bidirectional type system, provided that a conditional called \emph{mode-correctness} -- every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking --- is satisfied for every bidirectional typing rule.

Distinct from the classic Damas--Milner type inference, bidirectional type synthesis does \emph{not} require unification.
As noted by~\citet{Pierce2000}, annotations are propagated locally within adjacent nodes of a syntax tree, so introducing long-distance unification constraints in bidirectional type synthesis undermines its essence of locality.
Further, adding annotations for top-level definitions improves clarity, making the purpose of a program easier to understand, so it should be considered useful for the sake of documentation.
It is fair to say that bidirectional type synthesis is the modern foundation of type-checking technology across diverse languages.

Unlike parsing, nevertheless, which has a comprehensive theory and widely applicable practical techniques, bidirectional typing has only been developed on a case-by-case basis without a theory.
While it is easy to derive a bidirectional type synthesis algorithm, a general theory providing logical specifications with rigorously proven properties for a broad range of systems is still lacking.
As a result, we can only present design principles that we learned from individual systems loosely.
Moreover, unlike the plethora of available parser generators, `type-checker generators' rarely exist, so each type checker has to be independently built from scratch.

To tackle these challenges, this paper presents a theory of bidirectional typing and a verified implementation of bidirectional type synthesis for simple type theories.
There are many ways and extensions for designing bidirectional typing rules, such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}.
Yet, once the definition of a bidirectional type system is in place, deriving an algorithm becomes straightforward.
Accordingly, our theory's objective is not to formulate rules for various language features, but to introduce a formalism for bidirectional typing, analogous to grammar in parsing.
We introduce the notion of bidirectional binding signatures to specify bidirectional type systems and characterise essential criteria including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a type synthesis algorithm, as informally outlined by~\citet{Dunfield2021}.
For simplicity, we confine our discussion to bidirectional simple type systems that have \emph{syntax-directed} typing rules and leave the problem of presenting bidirectional typing for more advanced type theories in general as future work.

\subsection{Related work}\label{sec:related-work}
As we will formalise our theory for not only simple type theories but also in a manner distinct from typical formulations in bidirectional typing, it is pertinent to discuss related work upon which this work is built, apart from bidirectional typing, in order to articulate our contributions.

\subsubsection{Language formalisation and its frameworks} \label{sec:language-formalisation}
The vision of formalising the metatheory of every programming language was initiated by the \PoplMark challenge~\citep{Aydemir2005}.
This is has been exemplified in a textbook by~\citet{Wadler2022}, where language concepts are formally addressed using the proof assistant~\Agda, including bidirectional type synthesis for~\PCF.

@@

Earlier than the \PoplMark Challenge, \citet{Altenkirch1993} commented that rudimentary meta-operations and meta-properties for substitution constitute the bulk of language formalisation, motivating a number of language-formalisation frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to define the type of well-scoped/typed terms, substitution, term traversal and their meta-properties universally.
One of the core gadgets of these frameworks is the concept of \emph{descriptions} or \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying typing rules of a language.
It is noteworthy that these state-of-the-art frameworks cannot specify polymorphic type theories or dependent type theories yet and only \citeauthor{Allais2021} discussed meta-operations beyond substitution.

\subsubsection{Theories of abstract syntax with variable binding}\label{sec:theory-of-syntax}
The aforementioned frameworks except \citeauthor{Gheri2020}'s are at least inspired by \citet{Fiore1999}'s initial semantics for abstract syntax with variable binding using category theory.
The main idea is that the set of (untyped) abstract syntax trees for a language can be interpreted as the initial algebra for its signature endofunctor $\mathbb{\Sigma}$ on a family $\Term_{\Gamma}$ of well-scoped terms under the context~$\Gamma$ with
variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $X_{\Gamma}$ to $X_{\Delta}$, equivalently a free $\mathbb{\Sigma}$-algebra on the presheaf of variables $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$.
The endofunctor $\mathbb{\Sigma}$ encapsulates language constructs (except variables) while the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.

Constructing the initial algebra of terms boils down to defining an inductive type with a few constructors that align with the variable rule and the rule scheme for language constructs specified by a signature~\citep{Allais2021,Fiore2022}, without the need to explicitly define the colimit of a sequence of objects iteratively defined by the signature functor.

Substitution can also be modelled categorically, but it does not play a role in this paper, though.

@@

First, our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax} and is tailored towards \emph{extrinsically-typed}, \emph{syntax-directed} bidirectional type systems of simple types (\Cref{fig:raw-terms,fig:bidirectional-typing-derivations}) specified by a \emph{signature} for simple types (c.f.\ \Cref{re:type-signature}) and a \emph{bidirectional binding signature} for terms (\Cref{def:bidirectional-binding-signature}). 
The extrinsic typing is needed for type synthesis and our adherence to syntax-directed systems aligns with the standard idea of re-casting non-syntax-directed typing rules into their syntax-directed form to derive type synthesis algorithm~\citep{Peyton-Jones2007}.
As operational semantics are not needed for synthesising simple types, we do not address substitution.
Instead, our focus is on:
\begin{enumerate*}
  \item establishing the connection between the specified bidirectional type system and its simple type theory, i.e.\ \emph{soundness} and \emph{completeness} (\Cref{sec:soundness-and-completeness}); 
  \item formulating the \emph{mode-correctness} condition (\Cref{def:mode-correctness}) which suffices to establish the uniqueness of synthesised types (\Cref{thm:unique-syn}) and a decidable bidirectional type synthesis and checking (\Cref{thm:bidirectional-type-synthesis-checking}).
\end{enumerate*}

Our theory is based on Martin-L\"of type theory and formalised in \Agda (\Cref{sec:formalisation}), akin to those frameworks in \Cref{sec:language-formalisation}.
Moreover, we exploit the nature of constructive logic to formulate algorithmic soundness, completeness, and decidability. 
Recall that the law of excluded middle does not hold universally as an axiom in Martin-L\"of type theory, thus making a \emph{decidable statement} $P \vee \neg P$ non-trivial to prove.
Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
Further, suppose that we have a proof of the following statement:
\begin{quote}
  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some $A$ or any derivation of\, $\Gamma |- t : A$ for some $A$ leads to a contradiction'
\end{quote}
or rephrased succinctly as 
\begin{quote}
  `A context $\Gamma$ and a raw term $t$ \emph{decide} whether $\Gamma |- t : A$ has a derivation for some $A$',
\end{quote}
echoing the algorithm given by \citeauthor{Wadler2022} as noted in \Cref{re:type-synthesis-as-decidability-proof}.
Both algorithmic soundness and completeness are implied:
the proof either yields a typing derivation for the given raw term~$t$ or a proof that such a derivation is impossible where the former case is algorithmic soundness and the latter is the contrapositive form of completeness.
That is, our decidable bidirectional type synthesis (\Cref{thm:bidirectional-type-synthesis-checking}) provides algorithmic soundness, completeness, and decidability all at once.

In contrast to \citeauthor{Wadler2022}, our theory starts with raw terms and a general type synthesis problem, independent of bidirectional typing.
This starting point uncovers nuances regarding the lack of enough type annotations in bidirectional typing, often overlooked in literature, and thus distinguishes completeness (with respect to a type assignment system) from annotatability, concepts conflated by \citet{Dunfield2021}. 
To clarify the difference (\Cref{sec:annotatability}), we propose \emph{mode derivations} (\Cref{fig:mode-derivations,fig:generalised-mode-derivations}) and \emph{mode preprocessing} (\Cref{sec:mode-preprocessing}) assigning a mode to a raw term with enough annotations or otherwise pinpoints missing annotations.
By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we arrive our main result (\Cref{cor:trichotomy}) modelling a type synthesiser which checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has a typing derivation with respect to the specified simple type theory---a \emph{trichotomy} of raw terms.
Our bidirectional type synthesis can be instantiated for any system specified by a mode-correct signature, effectively a verified type-checker generator.

In summary, we contribute a theory of bidirectional type synthesis that is:
\begin{enumerate}
  \item \emph{simple yet general} for any system that can be specified by a bidirectional binding signature;
  \item \emph{constructive}, preferring
    \begin{enumerate*}
      \item logical decidability over algorithmic soundness, completeness, and decidability;
      \item mode preprocessing over annotatability.
    \end{enumerate*}
\end{enumerate}
Moreover, all these concepts have been formally developed in \Agda.

Plan of this paper:
\begin{enumerate}
  \item Introduction (\Cref{sec:intro}) 
    \LT{4 pp}
  \item Key ideas (\Cref{sec:key-ideas})
    \Josh{2.5 pp}
  \item Definitions for bidirectional type systems (bidirectional binding signature, bidirectional type systems, signature erasure and mode annotation) (\Cref{sec:defs})
    \LT{4 pp}
  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:pre-synthesis})
    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
    \Josh{3 pp}
  \item Bidirectional type inference (\Cref{sec:type-synthesis})
    \LT{4 pp}
  \item Formalisation and further examples (\Cref{sec:formalisation})
    \LT{5.5pp}
  \item Concluding remarks (\Cref{sec:future})
    \LT{1 p} \Josh{1p}
\end{enumerate}

>>>>

Type checking and inference serve as the transition to type-checked programs, also known as \emph{well-typed terms}, from parsed abstract syntax trees, referred as \emph{raw terms} by contrast.
Type inference algorithms were conceived to ascertain the type of any raw term without any type annotations.
However, it was later found that full parametric polymorphism leads to undecidability in type inference, as do dependent types~\citep{Wells1999,Dowek1993}.
In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing decidable algorithms for determining the types of \emph{suitably annotated} programs in languages with simple types, polymorphic types, dependent types, among others. 
\citet{Dunfield2021} summarised its design principles and applications that we know so far.

The idea of bidirectional type synthesis begins with extending typing judgements with two modes that direct the flow of types:
\begin{enumerate*}
  \item $\Gamma |- \isTerm{t} \syn A$ for synthesis and 
  \item $\Gamma |- \isTerm{t} \chk A$ for checking.
\end{enumerate*}
The former means that the type of a term is synthesised, using both the term and its context as input, while for the latter a term is to be checked against a given type.
Then, a type synthesis algorithm can be `read off' from a bidirectional type system, provided that a conditional called \emph{mode-correctness} -- every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking --- is satisfied for every bidirectional typing rule.

Contrary to the classic Damas--Milner type inference, bidirectional type synthesis does \emph{not} require unification.
\citet{Pierce2000} noted that bidirectional type synthesis propagates annotations locally within adjacent nodes of a syntax tree, so introducing long-distance unification constraints undermines the essence of locality in bidirectional type synthesis.
Further, annotations for, say, top-level definitions improve clarity, making the purpose of a program easier to understand, so they should not be considered redundant.
In light of these considerations, bidirectional type synthesis can be deemed as a technique that is more fundamental than unification and is a general design paradigm capable of handling a broad spectrum of programming languages.

Unlike parsing, nevertheless, which has a comprehensive theory and widely applicable practical techniques, bidirectional typing has only been developed on a case-by-case basis without a theory.
While it is straightforward to derive a type synthesis algorithm, a general theory providing logical specifications and rigorously proven properties for a class of systems is still lacking.
As a result, we can only present design principles that we learned from individual systems loosely.
Moreover, unlike the plethora of available parser generators, `type-checker generators' rarely exist, so each type checker, based on bidirectional typing or not, has to be independently built from scratch.

To tackle these challenges, this paper presents a theory of bidirectional typing and a verified implementation of bidirectional type synthesis for simple type theories.
There are many ways and extensions for designing bidirectional typing rules, such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}.
Yet, once the definition of a bidirectional type system is in place, deriving an algorithm becomes straightforward.
Accordingly, our theory's objective is not to formulate rules for various language features, but to introduce a formalism for bidirectional typing, analogous to grammar in parsing.
We introduce the notion of bidirectional binding signatures to specify bidirectional type systems and characterise essential criteria including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a type synthesis algorithm, as informally outlined by~\citet{Dunfield2021}.
For simplicity, we confine our discussion to bidirectional simple type systems that have \emph{syntax-directed} typing rules and leave the problem of presenting bidirectional typing for more general and advanced type theories as future work.

\subsection{Related work}\label{sec:related-work}
As our theory is developed formally not only for a class of simple type theories but also in a manner distinct from typical formulations in bidirectional typing, it is pertinent to discuss related work upon which this work is built, apart from bidirectional typing, in order to explain our contributions.

\subsubsection{Language formalisation and its frameworks} \label{sec:language-formalisation}
The vision of formalising the metatheory of every programming language was initiated by the \PoplMark challenge~\citep{Aydemir2005}.
This has been exemplified in the textbook by~\citet{Wadler2022} where language concepts are formally addressed using the proof assistant~\Agda, including bidirectional type synthesis for~\PCF.

@@

Earlier than the \PoplMark challenge, \citet{Altenkirch1993} commented that rudimentary meta-operations and their meta-properties constitute the bulk of formalisation, motivating a number of frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to define the type of well-scoped/typed terms, substitution, term traversal and their meta-properties universally.
One of the core gadgets of these frameworks is the concept of \emph{descriptions} or \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying typing rules of a language.
It is noteworthy that these state-of-the-art frameworks cannot specify polymorphic type theories or dependent type theories yet and only \citeauthor{Allais2021} discussed meta-operations beyond substitution.

\subsubsection{Theories of abstract syntax with variable binding}\label{sec:theory-of-syntax}
The aforementioned frameworks except \citeauthor{Gheri2020}'s are at least inspired by \citet{Fiore1999}'s initial semantics for abstract syntax with variable binding using category theory.
The main idea is that the set of (untyped) abstract syntax trees for a language consists of
\begin{enumerate*}
  \item a family $\Term_{\Gamma}$ of well-scoped terms under the context~$\Gamma$ with
  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $X_{\Gamma}$ to $X_{\Delta}$, thus a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, and
  \item an initial algebra structure given by the variable rule as a map from the embedding $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$ to $\Term$ and other language constructs as $\mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts, the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encapsulates language constructs (except variables), and the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
\end{enumerate*}
To put it succinctly, it is the free $\mathbb{\Sigma}$-algebra over the presheaf of variables $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$.

Fortunately, constructing the initial algebra of terms in type theory boils down to defining an inductive type with a few constructors that align with the variable rule and a rule scheme for language constructs specified by a signature~\citep{Allais2021,Fiore2022}.

Substitution is also modelled categorically, but it does not play a role in this paper, though.

@@

First, our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax} and is tailored towards \emph{extrinsically-typed}, \emph{syntax-directed} bidirectional type systems of simple types (\Cref{fig:raw-terms,fig:bidirectional-typing-derivations}) specified by a \emph{signature} for simple types and a \emph{bidirectional binding signature} for terms (\Cref{def:bidirectional-binding-signature} and c.f.\ \Cref{re:type-signature}). 
The extrinsic typing is needed for type synthesis and our adherence to syntax-directed systems aligns with the standard idea of re-casting non-syntax-directed typing rules into their syntax-directed form to derive type synthesis algorithm~\citep{Peyton-Jones2007}.
As substitution is not needed for synthesising simple types, it is not addressed.
Instead, our focus is on:
\begin{enumerate*}
  \item establishing the link between the specified bidirectional type system and its simple type theory, i.e.\ \emph{soundness} and \emph{completeness} (\Cref{sec:soundness-and-completeness}); 
  \item defining the \emph{mode-correctness} condition (\Cref{def:mode-correctness}) which suffices to establish the uniqueness of synthesised types (\Cref{thm:unique-syn}) and a decidable bidirectional type synthesis and checking (\Cref{thm:bidirectional-type-synthesis-checking}).
\end{enumerate*}

Our theory is based on Martin-L\"of type theory and formalised in \Agda (\Cref{sec:formalisation}), akin to those frameworks in \Cref{sec:language-formalisation}.
More importantly, we exploit the computational and logical aspects of type theory to formulate algorithmic soundness, completeness, and decidability. 
Recall that the law of excluded middle does not hold universally as an axiom in Martin-L\"of type theory, so a \emph{decidable statement} $P \vee \neg P$ is non-trivial.
Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
Further, suppose that we have a proof of the following statement:
\begin{quote}
  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some $A$ or any derivation of\, $\Gamma |- t : A$ for some $A$ leads to a contradiction'
\end{quote}
or rephrased succinctly as 
\begin{quote}
  `A context $\Gamma$ and a raw term $t$ \emph{decide} whether $\Gamma |- t : A$ has a derivation for some $A$',
\end{quote}
echoing the algorithm given by \citeauthor{Wadler2022} as noted in \Cref{re:type-synthesis-as-decidability-proof}.
Both algorithmic soundness and completeness are implied:
the proof either yields a typing derivation for the given raw term~$t$ or a proof that such a derivation is impossible where the former case is algorithmic soundness and the latter is algorithmic completeness in contrapositive form.
That is, our decidable bidirectional type synthesis (\Cref{thm:bidirectional-type-synthesis-checking}) provides algorithmic soundness, completeness, and decidability.

In contrast to \citeauthor{Wadler2022}, our theory starts with raw terms and a general type synthesis problem, independent of bidirectional typing.
This starting point uncovers nuances regarding the lack of enough type annotations in bidirectional typing, often overlooked in literature, and thus distinguishes completeness (with respect to a type assignment system) from annotatability, concepts conflated by \citet{Dunfield2021}. 
To clarify the difference (\Cref{sec:annotatability}), we propose \emph{mode derivations} (\Cref{fig:mode-derivations,fig:generalised-mode-derivations}) and \emph{mode preprocessing} (\Cref{sec:mode-preprocessing}) assigning a mode first to a raw term with enough annotations or otherwise pinpoints missing annotations.
By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result (\Cref{cor:trichotomy}) --- a \emph{trichotomy} of raw terms: a type synthesiser which checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has a typing derivation with respect to the specified simple type theory.
Our bidirectional type synthesis can be instantiated for any system specified by a mode-correct signature, effectively a verified type-checker generator.

In summary, we contribute a theory of bidirectional type synthesis that is:
\begin{enumerate}
  \item \emph{simple yet general} for any system that can be specified by a bidirectional binding signature;
  \item \emph{constructive}, based on Martin-L\"of theory and preferring logical decidability over algorithmic soundness, completeness, and decidability; mode preprocessing over annotatability.
\end{enumerate}
Moreover, all these concepts have been formally developed in \Agda.

This paper is structured in logical order as follows.
We first present an overview of our theory using simply typed $\lambda$-calculus in \Cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \Cref{sec:defs}.
Following this, we discuss the connection between a specified bidirectional type system and its associated simple type theory in \Cref{sec:pre-synthesis}.
In \Cref{sec:type-synthesis}, we introduce mode-correctness and bidirectional type synthesis.
We conclude by reflecting on related topics and potential challenges in extending our theory to a more general setting.

\todo[inline,caption={}]{
\begin{enumerate}
  \item Introduction (\Cref{sec:intro}) 
    \LT{4 pp}
  \item Key ideas (\Cref{sec:key-ideas})
    \Josh{2.5 pp}
  \item Definitions for bidirectional type systems (bidirectional binding signature, bidirectional type systems, signature erasure and mode annotation) (\Cref{sec:defs})
    \LT{4 pp}
  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:pre-synthesis})
    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
    \Josh{3 pp}
  \item Bidirectional type inference (\Cref{sec:type-synthesis})
    \LT{4 pp}
  \item Formalisation and further examples (\Cref{sec:formalisation})
    \LT{5.5pp}
  \item Concluding remarks (\Cref{sec:future})
    \LT{1 p} \Josh{1p}
\end{enumerate}
}
